---
title: "mini_project_1"
author: "Zhanchao Yang"
date: "2024-10-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(sf)
library(nngeo)
library(units)
library(pacman)
library(lubridate)
```
Load required data
```{r, include=FALSE}
station<- st_read("data/phl_bike_stations.shp")
add <- read_csv("data/indego-stations-2024-07-01.csv")
rider <- read_csv("data/indego-trips-2024-q2.csv")
neighborhood <- st_read("data/philadelphia-neighborhoods.shp")
```

## join the station with additional information csv
```{r}
station <- station %>%
  left_join(., add, by = c("id" = "Station_ID"))
```

## cleaning the rider table

```{r}
rider <- rider %>%
  select(duration, start_time, end_time, start_station, end_station, trip_route_category, bike_type)
```

## The 10 stations with the HIGHEST number of STARTS (stations that are origins)
```{r}
rider_starter <- rider %>%
  group_by(start_station) %>%
  summarize(n = n()) %>%
  arrange(desc(n))
```

```{r}
head(rider_starter, 10)
```
## The 10 stations with the LOWEST number of STARTS (stations that are origins)
```{r}
tail(rider_starter, 10)
```
## A map that displays the number of STARTS by station that shows the range of values
```{r}
start_station_join <- station %>%
  left_join(., rider_starter, by = c("id" = "start_station"))
```

## The 10 stations with the HIGHEST number of ENDS (stations that are destinations)
```{r}
rider_ender <- rider %>%
  group_by(end_station) %>%
  summarize(n = n()) %>%
  arrange(desc(n))
```

```{r}
head(rider_ender, 10)
```

## The 10 stations with the LOWEST number of ENDS (stations that are destinations)
```{r}
tail(rider_ender, 10)
```

## A map that displays the number of ENDS by station that shows the range of values
```{r}
end_station_join <- station %>%
  left_join(., rider_ender, by = c("id" = "end_station"))
```

## The 5 STARTING STATIONS with the longest AVERAGE trip durations
```{r}
average_duration_by_station <- rider %>%
  group_by(start_station) %>%
  summarize(average_duration = mean(duration, na.rm = TRUE))%>%
  arrange(desc(average_duration))
```

```{r}
head(average_duration_by_station, 5)
```


## the 5 STARTING stations with the lowest AVERAGE trip durations
```{r}
tail(average_duration_by_station, 5)
```

## What stations,BASED ON THE YEAR THEY WERE ACTIVATED, had the most trips?
```{r}
staion_ender_starter <- left_join(rider_ender, rider_starter, by = c("end_station" = "start_station")) %>%
  mutate(total = n.x + n.y) %>%
  select("end_station", "total")%>%
  rename("id" = "end_station")
```

```{r}
station_complete <- station %>%
  left_join(., staion_ender_starter, by = c("id" = "id"))
```

```{r}
station_complete <- station_complete %>%
  rename("Activation_date"="Day of Go_live_date")
```

```{r}
station_complete <- station_complete %>%
  mutate(Acti_date = mdy(Activation_date)) %>%
  mutate(year = year(Acti_date))
```

```{r}
station_trip_by_year <-station_complete %>%
  group_by(year,id) %>%
  summarize(total_trip = sum(total), .groups = "drop")
```

```{r}
top_stations_by_year <- station_trip_by_year %>%
  group_by(year) %>%
  filter(total_trip== max(total_trip))
```



## What are the 5 neighborhoods with the most TOTAL TRIPS and what are the 5 neighborhoods with the most TRIPS PER STATION?
```{r}
neighborhood <-neighborhood %>%
  st_transform(crs=st_crs(station_complete))
```
Check the invalid geometries in the neighborhood data
```{r}
invalid_neighborhoods <- neighborhood %>%
  filter(!st_is_valid(.))

# Print out how many invalid geometries there are
print(paste("Number of invalid geometries:", nrow(invalid_neighborhoods)))

neighborhood <- st_make_valid(neighborhood)
```
Spatial join the station data with the neighborhood data
```{r}
station_complete <- station_complete %>%
  filter(!is.na(total))
```

```{r}
station_in_neighborhood <- station_complete %>%
  st_join(neighborhood, join= st_within)
```

```{r}
station_in_neighborhood <- station_in_neighborhood %>%
  group_by(NAME) %>%
  summarize(total_trip = sum(total), total_station = n()) %>%
  mutate(trip_per_station = total_trip/total_station) %>%
  arrange(desc(total_trip))
```

## A categorical map that colors the stations by three categories for trip duration. You determine what values correspond to which class, but the map should have three categories: short, medium and long.
```{r}
average_duration_by_station <- average_duration_by_station %>%
  mutate(average_duration_cat = case_when(
    average_duration <= 20 ~ "short",
    average_duration > 20 & average_duration <= 30 ~ "medium",
    average_duration > 30 ~ "long"
  ))%>%
  rename("id" = "start_station")
```

```{r}
average_duration_staion_join <- left_join(station, average_duration_by_station, by = c("id" = "id"))
```


## get census data
```{r}
library(tidycensus)
```

```{r}
acs_2022 <- load_variables(2022, dataset="acs5")
acsVaraibles <- c(totpop = "B01001_001",
                  tot_commute="B08006_001",
                  bikers ="B08006_014",
                  median_age="B01002_001"
                  )
```

```{r,echo= TRUE}
Philiy_acs_2022 <- get_acs(geography="tract",
                           variables=acsVaraibles,
                           state="PA",
                           county="Philadelphia",
                           survey='acs5',
                           year=2022,
                           output="wide",
                           geometry=TRUE)
```


```{r}
Philiy_acs_2022 <- Philiy_acs_2022 %>%
  mutate(bike_rate = bikersE/tot_commuteE)
```

## Land use data munpulation
```{r}
land_use <- st_read("data/Land_Use.shp")
```
```{r}
land_use <- land_use %>%
  mutate(land_use = case_when(
    C_DIG1 == 1 ~ "Residential",
    C_DIG1 == 2 ~ "Commercial",
    C_DIG1 == 3 ~ "Industrial",
    C_DIG1 == 4 ~ "Civic/Institution",
    C_DIG1 == 5 ~ "Transportation",
    C_DIG1 == 6 ~ "Culture recreation",
    C_DIG1 == 7 ~ "Park Open Space",
    C_DIG1 == 8 ~ "Water",
    C_DIG1 == 9 ~ "Vacant or Other"
  ))
```

```{r}
land_use_e <- land_use %>%
  select(land_use)
```

```{r}
st_write(land_use_e, "land_use_e.gpkg")
```
